# The Guide {#guide}

1. Review [7 Best Practices](https://ahurford.github.io/quant-guide-all-courses/style.html) chapter of [Quantitative skills for biology](https://ahurford.github.io/quant-guide-all-courses/). Make a folder on your computer Desktop where you will work on this assignment.  

2. Download the file `wood.csv` from the course BrightSpace.

3. Review [8.2 Loading or importing data](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#loading-or-importing-data). Open and save a new R Script (see [3.2 Source Pane](https://ahurford.github.io/quant-guide-all-courses/rstudio.html#source-pane]) if you are unsure how to do this). One of the first commands of your R Script will be to load your data, for example:


```{r}
data1 <- read.csv(file = "woods.csv")
```

4. In the [Console](https://ahurford.github.io/quant-guide-all-courses/rstudio.html#console-pane), inspect your data to check that it has loaded. For example:

```{r}
head(data1)
```

If your data has not loaded, the most likely problem is a spelling error or problems with specifying the [path](https://ahurford.github.io/quant-guide-all-courses/rintro.html#working-directory) to `wood.csv`. You might try:

- the [RStudio way](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#the-rstudio-way-the-import-feature) of importing your data, or

- moving `wood.csv` to your [working directory](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#the-rstudio-way-the-import-feature).

5. You will want to do some exploration of your data. See [8.4 Handling the data](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#handling-the-data) for how to recover the names of the columns for all the data.

```{r}
names(data1)
```

See the metadata file on the course BrightSpace for an explanation of the meanings of the data columns. Read over the assignment instructions.

Do you hypothesize a relationship between the data in any of the columns? You might start investigating your hypothesis by making a plot, for example, showing the `Average cross sectional vessel area` and `the number of vessels`. To extract a column of the data see [8.4 Handling the data](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#handling-the-data).

For the `Average cross sectional vessel area`, in the _Console_, try:

```{r}
data1$A..mm.2.
```

Be sure to type exactly `A..mm.2.` after the `$`. You can make your plot as:

```{r}
plot(data1$N..mm..2.,data1$A..mm.2., pch = 19, xlab = "Average cross sectional vessel area", ylab = "Number of vessels per unit cross sectional area", col="blue")
```

Again, be sure to type the column names exactly as the appear in the output of `names(data1)` but without the `""`.

Can you make a plot of different variables, for example `Wood density` and `Average of absolute value of latitude`?

What happens if you type `plot(data1$Family, data1$Binomial)` in the `Console`? Why is this not a very sensible choice for something to plot?

To analyze the relation between variables we can use the `cor(x,y,method = "pearson")` function. And it will give us a numeric description of the data's relation (for more information about this topic:)


##    Sample with t.test {#t.test}
**Structural adaptation**: Average cross sectional vessel area

**Categories**:

  -   Tall(Average plant height is greater than 3 meters)

```{r}
cat1 <-data1[0:10,0:3]
cat1
```


  -   Short(Average plant height is less than 3 meters)

```{r}
cat2 <- data1[11:20,0:3]
cat2
```

  Means and standard deviations of both group of data:
  
```{r}
mean1 <- mean(cat1[,3])
mean2 <- mean(cat2[,3])

sd1 <- sd(cat1[,3])
sd2 <- sd(cat2[,3])

mean1
mean2
sd1
sd2
barplot(c(mean1,mean2), names.arg = c('cat1','cat2'),main ="Average of groups", ylim = c(0,0.0125), col = c('blue','blue'))
```

For the t.test if we want to evaluate $\alpha = 0.05$ if our p-value is greater than our $\alpha$ both means are not statistically different (based on the datasets). Althought, if the p-value is less than $\alpha$ then both means are statistically different: 

```{r}
t.test(cat1[,3], cat2[,3], alternative = 'two.sided',paired = TRUE)
```

For more information about t.tests and statistical analysis refer to:
<https://ahurford.github.io/quant-guide-all-courses/statistics.html>
