# The Guide {#guide}

To start solving this assignments what we want to do first is to load the data given
```{r}
data1 <- read.csv(file = "_B2010 2_Walk through Woods_DATASET and VALUES_POST.csv")
data1[0:5,0:9]
```

With this data we can make a plot to show correlation between the `Average cross sectional vessel area` and `the number of vessels` in the following way:

```{r}
plot(data1$N..mm..2.~data1$A..mm.2., pch = 19, xlab = "Average cross sectional vessel area", ylab = "Number of vessels per unit cross sectional area", col="blue")
```

And also we can look at the relation between `Wood density` and `Average of absolute value of latitude`
```{r}
plot(data1$AbsLat..Degrees.~data1$D..g.cm.3., pch = 19, xlab = "Wood density", ylab = "Average of absolute value of latitude", col = 'black')
```


To analyze the relation between variables we can use the `cor(x,y,method = "pearson")` function. And it will give us a numeric description of the data's relation (for more information about this topic:)


##    Sample with t.test {#t.test}
**Structural adaptation**: Average cross sectional vessel area

**Categories**:

  -   Tall(Average plant height is greater than 3 meters)

```{r}
cat1 <-data1[0:10,0:3]
cat1
```


  -   Short(Average plant height is less than 3 meters)

```{r}
cat2 <- data1[11:20,0:3]
cat2
```

  Means and standard deviations of both group of data:
  
```{r}
mean1 <- mean(cat1[,3])
mean2 <- mean(cat2[,3])

sd1 <- sd(cat1[,3])
sd2 <- sd(cat2[,3])

mean1
mean2
sd1
sd2
barplot(c(mean1,mean2), names.arg = c('cat1','cat2'),main ="Average of groups", ylim = c(0,0.0125), col = c('blue','blue'))
```

For the t.test if we want to evaluate $\alpha = 0.05$ if our p-value is greater than our $\alpha$ both means are not statistically different (based on the datasets). Althought, if the p-value is less than $\alpha$ then both means are statistically different: 

```{r}
t.test(cat1[,3], cat2[,3], alternative = 'two.sided',paired = TRUE)
```

For more information about t.tests and statistical analysis refer to:
<https://ahurford.github.io/quant-guide-all-courses/statistics.html>
