---
title: "A walk through the woods"
site: bookdown::bookdown_site
output: bookdown::gitbook #bookdown::html_document2
documentclass: book
github-repo: atellezcastr/walkThroughWoodsGuide
description: "Guide a to Walk through woods Assignment"
---

# Getting started and Learning R {-}

If you have not used `R` previously, we recommend that you work through Chapters 1 to 10 of the [Quantitative training guide](https://ahurford.github.io/quant-guide-all-courses/), which includes instructions on how to install `R` on a computer.

# How to do the analysis in R {-}

## Scatter plot and correlation {-}

Let's explore the _A walk through the woods_ data!

1.  It is best to make a folder on your computer Desktop where you will work on this assignment (if needed, review [Best Practices](https://ahurford.github.io/quant-guide-all-courses/style.html)).

2.  Download the file `woods.csv` (from the course BrightSpace) to the folder you made.

3.  [Open and save a new R Script](https://ahurford.github.io/quant-guide-all-courses/rstudio.html#source-pane%5D) your folder. One of the first commands of your R Script will be to [load your data](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#loading-or-importing-data). For example:

```{r}
data1 <- read.csv(file = "woods.csv")
```

4.  In the [Console](https://ahurford.github.io/quant-guide-all-courses/rstudio.html#console-pane), inspect your data to check that it has loaded. For example:

```{r, results=FALSE}
head(data1)
```

If your data has not loaded, the most likely problem is a spelling error or problems with specifying the [path](https://ahurford.github.io/quant-guide-all-courses/rintro.html#working-directory) to `woods.csv`. You might try:

-   the [RStudio way](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#the-rstudio-way-the-import-feature) of importing your data, or

-   moving `woods.csv` to your [working directory](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#the-rstudio-way-the-import-feature).

5.  After importing the data into R, read through the metadata file (below) to better understand the data.

> **Metadata Overview** Original Article Data used in this exerise originates from a Zanne, A. E., et al. (2010) study. The full article reference can be found below.
>
> **Columns in the data frame**
>
> Column 1: Family - The taxonomic family of the plant
>
> Column 2: Binomial - Refers to the genus and species for each plant
>
> Column 3: Amm - Average cross-sectional vessel area (in $mm^2$)
>
> Column 4: Nnn - Number of vessels per unit cross section area (in $mm^2$)
>
> Column 5: F - (F = $\frac{Amm^2}{Nmm^2}$ Fraction of cross sectional area that is in vessels (in $mm^2/mm^2$)
>
> Column 6: logA - log10 transformation of Amm (in $mm^2$)
>
> Column 7: logN - log10 transformation of Nmm (in $mm^2$)
>
> Column 8: logF - log10 transformation of F (in $\frac{Amm^2}{Nmm^2}$
>
> Column 9: Dens - Wood density measurement
>
> Column 10: AbsLat - Average of the absolute value of latitude (in degrees)
>
> **Reference**
>
> Zanne, A. E., Westoby, M., Falster, D. S., Ackerly, D. D., Loarie, S. R., Arnold, S. E. J., & Coomes, D. A. (2010). Angiosperm wood structure: Global patterns in vessel anatomy and their relation to wood density and potential conductivity. American Journal of Botany, 97(2), 207--215. <https://doi.org/10.3732/ajb.0900178>

```{r}
head(data1)
```

6.  You will want to do some exploration of your data. See [Handling the data](https://ahurford.github.io/quant-guide-all-courses/handling-data.html) for how to recover the names of the columns for all the data.

```{r}
names(data1)
```

7.  It is now helpful to think about the data given your knowledge of Botany. Do you hypothesize a relationship between the data in any of the columns? You might start investigating your hypotheses by making a plot, for example, showing the `Average cross-sectional vessel area` and `the number of vessels`. To extract a column of the data, see [Handling the data](https://ahurford.github.io/quant-guide-all-courses/handling-data.html). You may use either Approach 1: base R or Approach 2: `dplyr` (generally, the `dplyr` commands tend to be simpler, however, it is good to also know the base R commands).

**Approach 1: Base R**

For the `Average cross sectional vessel area`, in the `Console`, try:

```{r, results=FALSE}
data1$A.mm
```

Be sure to type exactly `A.mm` after the `$`. Try making your plot, including a line of best fit:

```{r}
plot(data1$A.mm, data1$N.mm, pch = 19, xlab = "Average cross sectional vessel area", ylab = "Number of vessels per unit cross sectional area", col="blue")
abline(lm(data1$N.mm ~ data1$A.mm))
```

You need to write the column names exactly as they appear in the output of `names(data1)` or RStudio will produce an error.

**Approach 2: dplyr**

To use the [`dplyr` functions](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#dplyr) you need to load the package (and [install](https://ahurford.github.io/quant-guide-all-courses/rintro.html#r-packages) if you have not already done so):

```{r,message=F}
library(dplyr)
```

For the `Average cross sectional vessel area`, in the `Console`, try:

```{r, results = F}
select(data1, A.mm)
```

Note that the output of `select(data1, A.mm)` is a data frame, but the plot function when used as `plot(x,y)` the `x` and `y` should be a vector. To make the same plot, but using `dplyr` functions, try:

```{r}
p = data.frame(select(data1, A.mm), select(data1, N.mm))
plot(p, pch = 19, xlab = "Average cross sectional vessel area",ylab = "Number of vessels per unit cross sectional area", col="blue")
abline(lm(p$N.mm ~ p$A.mm))
```

8.  Can you make a [plot](https://ahurford.github.io/quant-guide-all-courses/graph.html#scatter-plots-with-plotxy) of different variables, for example, _Wood density_ and _Average of the absolute value of latitude_?

9.  What happens if you type `plot(data1$Family, data1$Binomial)` in the `Console`? Why is this not a very sensible choice for something to plot?

10. To analyze the relationship between variables we can use the `cor(x,y,method = "pearson")` function. See [Correlation](https://ahurford.github.io/quant-guide-all-courses/statistics.html#correlation) for details. Can you find the correlation coefficient for any two reasonable variables in the `woods.csv` data set? Which two did you choose? If you plot these same two variables do you see a relationship?

## Comparing two means with a t.test {-}

This section will give a brief overview of how to perform a sample t-test to compare the average cross-sectional vessel area between two families: Rhamnaceae and Fabaceae. This example is to illustrate the coding for the analysis; you will want to use your knowledge of Botany to formulate a hypothesis to test.

1.  First, produce two variables that contain the data.

**Approach 1: Base R**
```{r, results=F}
Rhamnaceae = data1[data1$Family=="Rhamnaceae",]
Fabaceae = data1[data1$Family=="Fabaceae",]
```

Note that we are using `[]` to extract specific rows and columns of `data1`. Observe that `data1$Family=="Rhamnaceae"` means we want all the rows where the family is "Rhamnaceae", and the `,` indicates we would like all the columns. Type the following to understand more about these variables:

```{r, results=F}
head(Rhamnaceae)
head(Fabaceae)
length(Rhamnaceae[,1])
length(Fabaceae[,1])
```

The command `head()` reports the first 6 rows, and `length(Rhamnaceae[,1])` gives the length of column 1 of the variable `Rhamnaceae`.

**Approach 2: dplyr**

```{r, results=F}
Rhamnaceae = filter(data1,Family=="Rhamnaceae")
Fabaceae = filter(data1,Family=="Fabaceae")
```

You can also learn more about the data by using `length()` and `head()` as for the base R approach described above. To learn more about these functions see [here](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#base-r).

2.  We would like to perform a t-test to compare the means of the average cross-sectional vessel area for Rhamnaceae and Fabaceae, however, before we do this we will graph these data to observe if we expect the t-test to find a significant difference in means. An instructive graph to make is a boxplot, since this will allow us to visualize the median value for each of the two families, as well as other quantiles, and extreme values.

Using the `boxplot()` function requires the data to be arranged in a data frame with a specific organization:

**Approach 1: Base R**
```{r}
data = data.frame(Family = c(Rhamnaceae$Family,Fabaceae$Family), Area = c(Rhamnaceae$A.mm, Fabaceae$A.mm))
```

To see the resulting data frame, `data`, run these commands:
```{r}
head(data)
tail(data)
```

Note that `c()` binds lists together, and in the `Family` column of the new [data frame](https://ahurford.github.io/quant-guide-all-courses/rintro.html#data-structures) `data`, we need the value from the `$Family` column to be `Rhamnaceae` or  `Fabaceae`. In the `Area` column of `data` we need the average cross-sectional area of the vessels `$A.mm`. 

**Approach 2: dplyr**

In `dplyr`, we can make the data frame, `data`, by using [`select()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#selecting-a-column) to choose columns, and [`filter()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#filtering-data) to keep the families of interest. We can do this efficiently using [pipes](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#pipes):

```{r}
data = select(data1,Family, Area=A.mm) %>%
  filter(Family=="Rhamnaceae"|Family=="Fabaceae")
tail(data)
```

**Both approaches**

If we have made a data frame with the correct structure, then we can use the `boxplot()` command to visualize the average cross sectional area for vessels of plants from the families `Rhamnaceae` and  `Fabaceae`.

```{r, results=T}
boxplot(Area~Family, data, ylab = "Av. cross-sectional area of vessels (mm)")
```

Note that the rectangle for the boxplot is the interquartile range (from the 25th to 75th percentiles of our data). The 25th percentile means that only 25% of `Rhamnaceae` and  `Fabaceae` species have average cross-sectional vessel areas less than these values. From this box plot, it appears that the mean of the average cross-sectional vesseal area for Rhamnaceae species is different than for Fabaceae species, suggesting that we might expect the t-test to give a significant result.

4.  Performing a [t-test](https://ahurford.github.io/quant-guide-all-courses/statistics.html#t.test)) allows us to determine if the groups are statistically different and measures the probability that an observed difference could have occurred just by random chance. Here, we want to evaluate $\alpha = 0.05$, indicating that we are comfortable with a 5% risk of concluding a difference exists when there is no difference. If our p-value is greater than $\alpha$, both means are not statistically different.


```{r}
t.test(Rhamnaceae$A.mm, Fabaceae$A.mm, alternative = 'two.sided',paired = FALSE)
```

or, using `dplyr`:

```{r}
Rham.area = filter(data1,Family == "Rhamnaceae") %>% select(A.mm)
Faba.area = filter(data1,Family == "Fabaceae") %>% select(A.mm)
t.test(Rham.area, Faba.area)
```

Given the p-value of $1.76 \times 10^{-10}$, which is less than $\alpha = 0.05$, we reject our null hypothesis that the mean cross-sectional area of vessels is the same for Rhamnaceae and Fabaceae species. Therefore, our analysis suggest that the average cross-section vessel area for species of these Families are different.


